A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2020-04-10 paper/paper-revision.Rnw
Push edits with compiled doc (and working references)

In the United States, forensic analysis of shoe print evidence is varied and difficult. The process of collecting print evidence from a crime scene is not standardized, and it is often impossible to conduct an individualizing analysis (i.e., one based on randomly acquired characteristics) when a print is partial or of low quality. In fact, the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}. In the cases where individualizing characteristics are not used, it may be both possible and useful to assess the random match probability of the footwear evidence at the level of class characteristics; in other words, given a questioned shoe print, what is the probability that a randomly selected shoe in the relevant population has a similar tread pattern? Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns (e.g., knockoffs, similar models over different years, etc).

@@

However, a number of footwear studies (such as those by \citep{parent_sandy_significance_nodate} and \citep{benedict_geographical_2014}) have demonstrated that even a couple hundred shoe prints constitute an unweildy amount of data that require a large amount of time and effort to process and analyze. Thus, in parallel with the development of the aforementioned scanner, it is necessary to develop a way to process large amounts of data automatically. The primary focus of this paper is the development of one such method for large-scale automatic data processing.

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, our early prototypes of a footwear scanner have suggested that cameras are more robust than pressure scanners for placement in public locations, which may experience inclement weather or other non-controlled conditions. Thus, we are working to develop a scanner that uses cameras to take pictures of shoe outsoles, much like images that are currently found on the websites of online shoe retailers \autoref{fig:prototype-scanner}. Despite the fact that images from the scanner would be more degraded in some ways than images found from online retailers, the fundamental requirements to analyze such data would be similar. Thus, retailer images provide a reasonable basis for the development of a method to assess similarity in tread images. More specifically, this assessment is done by identifying a set of relevant features and using image analysis methods to detect the presence or absence of the features in each image, and then defining similarity measures based on those features to determine how many shoes have a tread pattern similar to the pattern of interest.

@@

\mt{A reasonable critique of this approach is that images collected from online retailers are of much higher quality than would be expected from the sidewalk scanner proposed in \autoref{sec:Introduction}, and thus the CNN described in this paper is not immediately ready to process real degraded images. This concern is especially fair when considering that machine learning methods tend to perform poorly when applied to test data that do not resemble the training data. However, the proposed automation methods for this approach are quite novel in the domain of footwear examination, and a fully functioning scanner does not yet exist to collect large amounts of degraded images. Thus, the goal of this work is to evaluate the proposed classification scheme and determine whether CNNS can sufficiently distinguish features found on outsoles. Addressing these questions with clean retailer images provides both justification and guidance for modifying this approach for real degraded images.}

>>>>

In the United States, forensic analysis of shoe print evidence is varied and difficult. The process of collecting print evidence from a crime scene is not standardized, and it is often impossible to conduct an individualizing analysis (i.e., one based on randomly acquired characteristics) when a print is partial or of low quality. In fact, the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}. In the cases where individualizing characteristics are not used, it may be both possible and useful to assess the random match probability of the footwear evidence at the level of class characteristics; in other words, given a questioned shoe print, what is the probability that a randomly selected shoe in the relevant population has a similar tread pattern? Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns (e.g., knockoffs, similar models over different years, etc). \fix{[This is an important point. Should I make a bigger deal out of it throughout the paper?]}

@@

A number of footwear studies (such as those by \citep{parent_sandy_significance_nodate} and \citep{benedict_geographical_2014}) have demonstrated that even a couple hundred shoe prints constitute an unweildy amount of data, which then require a large amount of time and effort to process and analyze. Thus, in parallel with the development of the aforementioned scanner, it is necessary to develop a way to process large amounts of data automatically. The primary focus of this paper is the development of one such method for large-scale automatic data processing.

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, our early prototypes of a footwear scanner have suggested that cameras are more robust than pressure scanners for placement in public locations, which may experience inclement weather or other non-controlled conditions. Thus, we are working to develop a scanner that uses cameras to take pictures of shoe outsoles, much like images that are currently found on the websites of online shoe retailers \autoref{fig:prototype-scanner}. Despite the fact that images from the scanner would be more degraded in some ways than images found from online retailers, the fundamental requirements to analyze such data would be similar, so retailer images provide a reasonable basis for the development of a method to assess similarity in tread images. More specifically, this assessment is done by identifying a set of relevant features and using image analysis methods to detect the presence or absence of the features in each image, and then defining similarity measures based on those features to determine how many shoes have a tread pattern similar to the pattern of interest.

@@

\mt{A reasonable critique of this approach is that images collected from online retailers are of much higher quality than would be expected from the sidewalk scanner proposed in \autoref{sec:Introduction}, and thus the CNN described in this paper is not immediately ready to process real degraded images. This concern is especially fair when considering that machine learning methods tend to perform poorly when applied to test data that do not resemble the training data. However, the proposed automation methods for this approach are quite novel within the domain of footwear examination, and a fully functioning scanner does not yet exist to collect large amounts of degraded images. Thus, the goal of this work is to evaluate the proposed classification scheme and determine whether CNNs can sufficiently distinguish features found on outsoles. Addressing these questions with clean retailer images provides both justification for further developing the scanner and guidance for modifying the model to process real degraded images.}
